---
title: "basicWorking"
author: "MeiSanderson"
date: "2026-04-29"
output: html_document
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
## Global options
library(pacman)

pacman::p_load(rmdformats, knitr)

options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


```{r load_libraries}

pacman::p_load(sfnetworks,
               sf,
               tidygraph,
               tidyverse,
               osmdata,
               raster,
               units,
               mapview)

pacman::p_load(slopes)


```



## Data Preparation

Loading, CRS conversion etc.

1)
```{r, prepare base map: aarhus commune}

## BASEMAP;   inspiration from class scripts

# get municipalities
muns <- readRDS("data/gadm36_DNK_2_sp.rds") %>% st_as_sf() 
muns$NAME_2[31] <- "Aarhus" # from århus to aarhus


# filter to only extract Aarhus
baseMap <- filter(muns, NAME_2 == "Aarhus")


# save CRS
baseMap_crs = st_crs(baseMap) # longlat, WGS84

# view
mapview(baseMap)


```


2)

```{r, prepare road network}

## inspiration from class scripts (specifically, class 06, which is based on luukvdmeer.github.io/UseR2024/notebook.html)


# query the stress within our area (i.e., Aarhus commune) from OpenStreetMap
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet)) # telling R that we do in fact have wifi

streets <- osmdata::opq(sf::st_bbox(baseMap)) %>% # generate surrounding box
  osmdata::add_osm_feature(key = "highway") %>% # highway describes all roads and footpaths; one could also set a value to get a specific type of road (e.g., living_street, pedestrian or motorway; wiki.openstreetmap.org/wiki/Map_features#Highway) - might be interesting for future research to assess which roads (are close to benches) and how much noise pollution come from them
  osmdata::osmdata_sf() %>%
  osmdata::osm_poly2line() %>%
  purrr::pluck("osm_lines") %>%
  sf::st_intersection(baseMap) %>%
  sf::st_cast("MULTILINESTRING") %>%
  sf::st_cast("LINESTRING")


# save the road data
saveRDS(streets, "data/streets_osm.rds")


```

```{r, load in the streets/road data if already computed}

streets <- readRDS("data/streets_osm.rds")

streets <- streets %>% 
  dplyr::select(osm_id, highway) # simplying the dataset

```



3)

```{r, prepare benches data}

## inspiration from luukvdmeer.github.io/UseR2024/notebook.html


# query the benches within our area (i.e., Aarhus commune) from OpenStreetMap
benches <- osmdata::opq(sf::st_bbox(baseMap)) %>% # generate surrounding box
  osmdata::add_osm_feature(key = "amenity", value = "bench") %>%
  osmdata::osmdata_sf() %>%
  purrr::pluck("osm_points") %>%
  sf::st_intersection(baseMap)


# save the road data
saveRDS(benches, "data/benches_osm.rds")

```

```{r, load in the benches data if already computed}

benches <- readRDS("data/benches_osm.rds")

print(colnames(benches)) # it includes bench data as well as various other details such as the colour, ability to lie down on it, separated seats etc.

```


_NOTE:_ As OpenStreetMap also uses the WGS84 (4326) CRS, there is no need for CRS conversion


```{r, visual assessment}

par(mar = c(0, 0, 0, 0))
plot(sf::st_geometry(streets))
plot(sf::st_geometry(benches), col = "red", pch = 20, cex = 0.5, add = TRUE)

```



## Combine (i.e., overlay) the layers

-> not needed?

```{r}

```





## Possible: Visualisation

We could consider plotting walking travel time to a bench for this map of all of Aarhus Commune - especially, if we are to focus on a smaller area of Aarhus Commune for the remainding analysis
-> still i think this might be overly simplistic

```{r}

## inspiration from class scripts (specifically, class 06, which is based on luukvdmeer.github.io/UseR2024/notebook.html)

# prepare the streets data
streets <- readRDS("data/streets_osm.rds")
streets <- streets %>% dplyr::select(osm_id, name, surface) # simply to work with less data at the moment

streets$length = sf::st_length(streets) %>% units::set_units("m")


# utilise a Digital Elevation Model (i.e., DEM), a spatial grid with elevation values for a certain area
dem <- raster("data/DNK_msk_alt.grd")
dem <- crop(dem, st_bbox(baseMap))
gradients <- slopes::slope_raster(streets, dem, directed = TRUE) * 100

st_crs(dem) # appears to be the same as streets


# set an average speed that can in- or decrease based on the gradient/slope of the street; though with a min and max speed
# CHOSEN PARAMETERS (as meters/hour): max speed = x, min speed = x; average speed (i.e., default) = x
# these are randomly chosen cause idk, and i actually don't think these plots are useful / meaningful
maxSpeed_walk = 20
minSpeed_walk = 5
avgSpeed_walk = 10

# !! if using a basic linear relation between gradient and speed
estimate_speed <-function(gradient, default = avgSpeed_walk, max = maxSpeed_walk, min = minSpeed_walk) {
  ifelse(  # is gradient less than 0 (i.e., going downhill)? then increase speed; if above 0, tehn decrease speed
    gradient < 0, 
    min(default + 0.8 * abs(gradient), max), # downhill
    max(default - 1.4 * gradient, min) # uphill
  )
}

streets$speed <- sapply(gradients, estimate_speed) %>% units::set_units("m/h")

plot(dplyr::select(streets, speed))


# convert to meters
streets$time <-units::set_units(streets$length / streets$speed, "min")

plot(dplyr::select(streets, time))


```





## Possible Next Step:  plot homeless shelters in Aarhus and assess the distance from those to benches
OpenStreetMap does have a value for the amenity key called 'social_facility', which describes facilities providing social services such as group & nursing homes etc. in addition to someless shelters. And idk if we can extract something specific from a value (like social_facility), so we might have to find the data elsewhere or plot it ourselves (assuming that there would only be so mnay homeless shelters).

-> i assume we are thinking specifically of places where sleep is possible(?); otherwise there are also various 'væresteder' but i don't think they give beds for the night

_examples found of homeless shelters in Aarhus (DK: herberger (specifically the sleeping arrangements) / varmestue ("a place to go") / hjemløsetilbud)_
- Aarhus Vinterherberg (specifically for non-Danish homeless people AND only in the winter months) (blaakors.dk/vores-arbejde/noedherberger)
- Varmestue Herning (in Herning, so only important if examining Aarhus Commune)
- Spor 7 (Godsbanen; natvarmestue)
- Kirkens Korshær Herberget Tre Ege
- Omsorgscenter Pitstop (viby; omsorgscenter for  syge hjemløse og funktionelt hjemløse i Aarhus Kommune) (rodekors.dk/vores-arbejde/socialt-arbejde/hjemloese/omsorgscenter-pitstop?gad_source=1&gad_campaignid=23814314290&gclid=CjwKCAjw5NvPBhAoEiwA_2egfiD_SeISgOoQMtUX1mCNiw08K9-zdqpzBCMfNZgBMhWqXyEHQQ19wxoCEQ4QAvD_BwE)


_potentially relevant sources_
- information about 'Varmestuer' in DK by Kirkens Korshær: https://kirkenskorshaer.dk/2025/09/varmestuer/
- 'når menneskesynet vakler: Aarhus svigter de mest udsatte': https://jyllands-posten.dk/jpaarhus/debat/breve/ECE18502465/naar-menneskesynet-vakler-aarhus-svigter-de-mest-udsatte/


 - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - - 
 
# Hot and Cold Spot Analysis


Using the sfhotspot library (https://pkgs.lesscrime.info/sfhotspot/)
- Introduction to library as used as inspiration (https://lesscrime.info/post/sfhotspot-1-0-0/)


-> as it mainly follows these workflows due look the functions up and assess


## Libraries
Not sure which is needed yet other than sfhotspot and ggplot2

```{r}

pacman::p_load(ggspatial, ggfortify) # ?
  
pacman::p_load(sfhotspot, # for the hot/cold-spot analysis
               ggplot2) # for nice plots


```


## Plot the basemap with our layers (just benches - for now at least)

text

```{r}

# haven't found the right functions


```


## 1) Counting points (i.e., number of benches)

-> important issues with this approach mentioned here: https://lesscrime.info/post/sfhotspot-1-0-0/

```{r}

benches <- readRDS("data/benches_osm.rds")

baseMap_grid <- hotspot_grid(baseMap, quiet = TRUE) # overlay grid over baseMap

benches_count_grid <- hotspot_count(benches, grid = baseMap_grid) # count benches pr. grid cell

ggplot() +
  autolayer(benches_count_grid, alpha = 0.75) +
  scale_fill_distiller(direction = 1) +
  #geom_sf(data = benches, size = 0.1) +  # plots the actual benchs
  labs(fill = "count of\npoints in\neach cell") +
  theme_void()


```


## 1.2) Estimating point density; alternative to the counts

Based on 2-dimensional kernel-densit-estimation using point densities instead of counts

```{r, with degree-based cells}

benches <- readRDS("data/benches_osm.rds")

baseMap_grid <- hotspot_grid(baseMap) # overlay grid over baseMap

benches_kde <- hotspot_kde(benches, grid = baseMap_grid) # try with different reductions (%) of the default bandwidth: bandwidth_adjust = 0.4

ggplot() +
  autolayer(benches_kde, alpha = 0.75) +
  geom_sf(data = baseMap, colour = "grey20", fill = NA) +
  scale_fill_distiller(direction = 1) +
  labs(title = "Density of Benches in Aarhus Commune", fill = "density of benches") +
  theme_void() 


```


```{r, with meter-based cells}


# transforming data from lat/lon to a suitable meter-based coordinate system for this plot to have cells of more meaningful (as least to us) sizes
benches <- readRDS("data/benches_osm.rds")

benches_utm <- st_transform_auto(benches) # automatically transform from lat/lon to a suitable meter-based system
baseMap_utm <- st_transform_auto(baseMap, quiet = TRUE) # quiet simply suppresses some messages
baseMap_grid_utm <- st_transform_auto(baseMap_grid, quiet = TRUE)


# actual plot

benches <- readRDS("data/benches_osm.rds")

baseMap_grid <- hotspot_grid(baseMap, quiet = TRUE) # overlay grid over baseMap

benches_kde_utm <- hotspot_kde(benches_utm, grid = baseMap_grid_utm, bandwidth_adjust = 0.4) # try with different reductions (%) of the default bandwidth

ggplot() +
  autolayer(benches_kde_utm, alpha = 0.75) +
  geom_sf(data = baseMap_utm, colour = "grey20", fill = NA) +
  scale_fill_distiller(direction = 1) +
  labs(title = "Density of Benches in Aarhus Commune", fill = "density of benches") +
  theme_void() 


```




## 2) Finding hotspots

Which cells in our grid have more benches than would be expected by "chance"

```{r}

benches_gistar <- hotspot_gistar(benches_utm, grid = baseMap_grid_utm, bandwidth_adjust = 0.4) %>%  # try with different reductions (%) of the default bandwidth
  dplyr::filter(gistar > 0, pvalue < 0.05)


# hotspots as per cells with significantly more benches tha expected per "change"
ggplot() +
  geom_sf(aes(color = kde, fill = kde), data = benches_gistar, alpha = 0.75, colour = NA) +
  geom_sf(data = baseMap_utm, colour = "grey20", fill = NA) +
  scale_fill_distiller(aesthetics = c("colour", "fill"), direction = 1) +
  labs(title = "Density of Benches in Aarhus Commune", fill = "density of benches (kde)") +
  theme_void() 


```


# OTHER POSSIBLE/POTENTIAL EXTENSIONS OF THE HOT/COLD SPOT ANALYSIS
- using 'hotspot_dual_kde()' to compare the kernel density of two layers of points;       perhaps to estimate the bench placements relative to local population or something




## Attribute analysis

Creating a foundation for classifying benches as inclusive vs exclusionary based on presence / absence of particular attributes.

```{r}

# benches dataset, as loaded above

# inclusive exclusionary - poissibly relevant attributes to consider
inex_cols <- c("backrest","armrest","lying_down","seats","covered","material", "seats.separated","two_sided")
#inex_cols

# simply for inspection
benches %>% 
  st_drop_geometry() %>% 
  dplyr::select(inex_cols) %>%
  summary()

# length is quite uninformative as yes... all the columns have an equal number of rows but NAs are
# overlooked so we attend to that next!

## count NA's in each inex_col

```


```{r} 

# now we create some basis for classifying the benches into inclusive vs exclusionary groups

inex_benches <- benches %>% 
  dplyr::mutate(
  
    # backrest
    has_backrest = case_when(
      backrest == "yes" ~ TRUE,
      backrest == "no"  ~ FALSE,
      TRUE              ~ NA   # anything else (including NA) stays unknown
    ),
    
    # armrest
    has_armrest = case_when(
      armrest == "yes" ~ TRUE,
      armrest == "no"  ~ FALSE,
      TRUE             ~ NA
    ),
    
    # can lie down
    can_lie_down = case_when(
      lying_down == "yes" ~ TRUE,
      lying_down == "no"  ~ FALSE,
      TRUE                ~ NA
    ),
    
    # covered (shelter from rain)
    is_covered = case_when(
      covered == "yes" ~ TRUE,
      covered == "no"  ~ FALSE,
      TRUE             ~ NA
    ),
    
    # separated seats (armrests between seats — exclusionary for rough sleeping)
    has_separated_seats = case_when(
      seats.separated == "yes" ~ TRUE,
      seats.separated == "no"  ~ FALSE,
      TRUE                     ~ NA
    ),
  
    
   # determining the state of a bench!
    is_inclusive = case_when(
      
    # --- exclusionary ---
    has_separated_seats == TRUE                                          ~ "exclusive",
    has_backrest == FALSE                                                ~ "exclusive",
    has_armrest == FALSE                                                 ~ "exclusive",
  
    # --- inclusive ---
    has_backrest == TRUE & has_armrest == TRUE & is_covered == TRUE      ~ "inclusive_covered",
    has_backrest == TRUE & has_armrest == TRUE                           ~ "inclusive",
  
    TRUE                                                                 ~ NA_character_
  ))

cat("Inclusivity classification:\n")
table(inex_benches$is_inclusive, useNA = "always")

```


## Map confirmed inclusive / exclusive / unknown

Actually not a helpful visual atm, i think i messed something up here with the var names so I'll return to it tomorrow.

```{r}

benches_utm <- st_transform_auto(inex_benches)

# separate into confirmed groups only — NA benches plotted separately
benches_in <- filter(benches_utm, is_inclusive == TRUE)
benches_ex <- filter(benches_utm, is_inclusive == FALSE)
benches_unknown   <- filter(benches_utm, is.na(is_inclusive))

ggplot() +
  geom_sf(data = baseMap_utm, fill = "grey97", colour = "grey60") +
  geom_sf(data = benches_unknown,   colour = "grey75", size = 0.5, alpha = 0.5) +
  geom_sf(data = benches_in, colour = "#2166ac", size = 3, alpha = 0.9) +
  geom_sf(data = benches_ex, colour = "#d6604d", size = 3, alpha = 0.9) +
  labs(
    title    = "Bench Inclusivity in Aarhus Commune",
    subtitle = "Blue = inclusive | Red = exclusive | Grey = unconfirmed (NA)"
  ) +
  theme_void()


```

## Voronoi - Service Areas (zoom to study area(s))

Need to adjust bbox coordinates for our chosen area - this is incredibly large and uninformative right now.
Can use mapview(baseMap_utm) to find the (approximately) right values interactively. 

Essentially, this section is just kind of a proof-of-concept right now!

```{r, voronoi study area}

study_bbox <- st_bbox(
  c(xmin = 565000, ymin = 6220000, xmax = 572000, ymax = 6226000),  # <- adjust
  crs = st_crs(baseMap_utm)
) %>% st_as_sfc()

benches_study_all <- st_intersection(benches_utm, study_bbox)
benches_study_in  <- st_intersection(benches_in, study_bbox)
benches_study_ex  <- st_intersection(benches_ex, study_bbox)
baseMap_study     <- st_intersection(baseMap_utm, study_bbox)

cat(sprintf("Benches in study area: %d total | %d inclusive | %d exclusive | %d NA\n",
  nrow(benches_study_all),
  nrow(benches_study_in),
  nrow(benches_study_ex),
  nrow(benches_study_all) - nrow(benches_study_in) - nrow(benches_study_ex)
))

```

```{r, voronoi poygons}

make_voronoi <- function(points, clip_area) {
  points %>%
    st_geometry() %>%
    st_union() %>%
    st_voronoi() %>%
    st_collection_extract("POLYGON") %>%
    st_sf() %>%
    st_intersection(clip_area) %>%
    mutate(area_m2 = as.numeric(st_area(geometry)))
}

voronoi_all <- make_voronoi(benches_study_all, study_bbox)
voronoi_in <- make_voronoi(benches_study_in, study_bbox)

```


```{r, voronoi plot}

p_vor_all <- ggplot() +
  geom_sf(data = voronoi_all, aes(fill = area_m2), colour = "white", linewidth = 0.2) +
  geom_sf(data = benches_study_all, size = 0.8, colour = "black") +
  geom_sf(data = baseMap_study, fill = NA, colour = "grey30") +
  scale_fill_viridis_c(option = "plasma", direction = -1, labels = scales::comma) +
  labs(title = "Voronoi: All Benches", fill = "service area (m²)") +
  theme_void()

p_vor_incl <- ggplot() +
  geom_sf(data = voronoi_in, aes(fill = area_m2), colour = "white", linewidth = 0.2) +
  geom_sf(data = benches_study_in, size = 0.8, colour = "#2166ac") +
  geom_sf(data = baseMap_study, fill = NA, colour = "grey30") +
  scale_fill_viridis_c(option = "mako", direction = -1, labels = scales::comma) +
  labs(title = "Voronoi: Inclusive Benches Only", fill = "service area (m²)") +
  theme_void()

#p_vor_all + p_vor_incl


```


```{r, voronoi summary}

cat("=== All benches — service area (m²) ===\n")
summary(voronoi_all$area_m2)

cat("\n=== Inclusive benches only — service area (m²) ===\n")
summary(voronoi_in$area_m2)

ratio <- mean(voronoi_in$area_m2) / mean(voronoi_all$area_m2)
cat(sprintf("\nMean service area is %.1fx larger when restricted to inclusive benches.\n", ratio))

```


## Demographic Overlay

An idea for using dkstat data which is a part of the opengov dk intitative: https://github.com/rOpenGov/dkstat 
We could use IDs to identify different demographics spread through aarhus - not sure if an ID for e.g., 'homeless' or socialt udsatte exists or one for elderly people exists but it might be a nice (maybe) way to get an understanding of how the presence and distribution of benches say something about how we care (or not) about various demographics of people.

```{r, elderly}

install.packages(
  "dkstat",
  repos = c(
    ropengov = "https://ropengov.r-universe.dev",
    getOption("repos")
  )
)

meta <- dkstat::dst_meta("", lang = "en") # need to test an ID
meta$variables

```


```{r, homeless}

# ...

```


