---
title: "basicWorking"
course: "Cultural Datascience: Spatial Analytics at AU"
authors: "Aiswarya Roy & Mie Norre Engemann"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
---

_Script Description:_
INSERT



_Script:_

```{r setup, include=FALSE}
## Global options
library(pacman)

pacman::p_load(rmdformats, knitr)

options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r}

knitr::clean_cache()
chooseBasemap <- "ringvej" # options: ringvej, ringgade, municipality

```


## Loading Libraries

```{r load_libraries, message = FALSE, warning=FALSE}
library(pacman)

pacman::p_load(sfnetworks,
               sf,
               tidygraph,
               tidyverse,
               osmdata,
               raster,
               units,
               mapview,
               ggplot2,
               geodata,
               dplyr,
               sfhotspot,
               geopackage)

# check which ones we actually use

```

```{r citation_stuff_deleteLater}
#packageVersion("geodata")
#citation(package="geodata")
#citation(package="osmdata")

#print(ourDownloadedData)

#print(ourDownloadedData_sf)
```


## Data Preparation (ME)

Loading Data, CRS conversion etc.

1) Base Map of Aarhus
```{r download_municipality_data}

## BASEMAP;   inspiration from class scripts
workingDirectory_path <- getwd()

# get municipalities
municipalityData_downloaded <- geodata::gadm(country="Denmark", level=2, path=file.path(workingDirectory_path, "data"), version="3.6") # getting data from the GADM database and save to data/gadm/
municipalityData_sf <- st_as_sf(municipalityData_downloaded)


```

```{r prepare_municipality_base_map}

## BASEMAP;   inspiration from class scripts

# preparing the data for this project
municipalityData_sf$NAME_2[31] <- "Aarhus" # from århus to aarhus

municipalityData_sf <- subset(municipalityData_sf, select = -c(GID_1, NL_NAME_1, GID_2, VARNAME_2, NL_NAME_2, CC_2)) # removing columns not used


# filter to only extract Aarhus
aarhus_municipality <- filter(municipalityData_sf, NAME_2 == "Aarhus")


# save CRS
baseMap_crs = st_crs(aarhus_municipality) # longlat, WGS84

# view
mapview(aarhus_municipality)


```


1.1) Ringvejen / Ringgaden
```{r}

# load the shapefile for noise pollution from streets during the night at 1.5 meters height
aarhus_ring <- read_sf("data/ringVejGade/ringVejGade.geojson")

aarhus_ringvej <- filter(aarhus_ring, Ringvejen == 1)
aarhus_ringgade <- filter(aarhus_ring, Ringgaden == 1)

mapview(aarhus_ringvej)

```





2) Bench Data

```{r currentMap}
chooseBasemap <- "ringvej"
if (chooseBasemap == "ringvej"){
  baseMap <- aarhus_ringvej
  
  benchRemoveCol <- c("artist_name", "artwork_type", "bin", "colour", "direction", "material", "note", "panoramax", "survey.date", "tourism")
  
} else if (chooseBasemap == "ringgade"){
  baseMap <- aarhus_ringgade
  benchRemoveCol <- c("artist_name", "artwork_type", "bin", "colour", "direction", "material", "note", "panoramax", "survey.date", "tourism")
  
} else if(chooseBasemap == "municipality") {
  baseMap <- aarhus_municipality
  benchRemoveCol <- c("artist_name", "artwork_type", "bin", "colour", "inscription", "not.inscription", "direction", "lit", "material", "not.tourism.artwork", "note", "panoramax", "source", "survey.date", "tourism", "level")
} else{
    stop("Input ringvej, ringgade or municipality for the chooseBasemap variable")
  }


cat("Basemap:", chooseBasemap, "\n")
print(st_bbox(baseMap))
print(st_geometry_type(baseMap))
mapview(baseMap)

```


```{r download_benches_data}

## inspiration from class scripts (specifically, class 06, which is based on luukvdmeer.github.io/UseR2024/notebook.html)


# query the roads within our area (i.e., Aarhus commune) from OpenStreetMap
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet)) # telling R that we do in fact have wifi

# query the benches within our area (i.e., Aarhus commune) from OpenStreetMap
benches_aarhus <- osmdata::opq(sf::st_bbox(baseMap)) %>% # generate bounding box
  osmdata::add_osm_feature(key = "amenity", value = "bench") %>%
  osmdata::osmdata_sf() %>%
  purrr::pluck("osm_points") %>%
  sf::st_intersection(baseMap) # fitting to Aarhus municipality


# save the road data
file.remove("data/benches_osm.rds")
bench_file <- sprintf("data/benches_osm_%s.rds", chooseBasemap)
saveRDS(benches_aarhus, bench_file)


cat("Benches:", nrow(benches_aarhus), "\n")
mapview(baseMap) + mapview(benches_aarhus)



```


```{r load_in_the_benches_data_if_already_computed}

benches <- readRDS(bench_file)

```


```{r removing_columns}


benches <- dplyr::select(benches, -all_of(benchRemoveCol)) # 'lit', 'material' and other features could be incorporated into the classifications of hostile, non-hostile and sleep-friendly benches but wasn't for the current project

mapview(baseMap) + mapview(benches)

```


_NOTE:_ As OpenStreetMap also uses the WGS84 (4326) CRS, there is no need for CRS conversion




```{r classify_benches} 

## CREATED BY AR

# now we create some basis for classifying the benches into hostile, non-hostile and sleep-friendly groups

inex_benches <- benches %>% 
  dplyr::mutate(
  
    # backrest
    has_backrest = case_when(
      backrest == "yes" ~ TRUE,
      backrest == "no"  ~ FALSE,
      TRUE              ~ NA   # anything else (including NA) stays unknown
    ),
    
    # armrest
    has_armrest = case_when(
      armrest == "yes" ~ TRUE,
      armrest == "no"  ~ FALSE,
      TRUE             ~ NA
    ),
    
    # can lie down
    can_lie_down = case_when(
      lying_down == "yes" ~ TRUE,
      lying_down == "no"  ~ FALSE,
      TRUE                ~ NA
    ),
    
    # covered (shelter from rain)
    if(chooseBasemap == "municipality"){
      is_covered = case_when(
        covered == "yes" ~ TRUE,
        covered == "no"  ~ FALSE,
        TRUE             ~ NA)
    },
    
    # separated seats (armrests between seats — exclusionary for rough sleeping)
    has_separated_seats = case_when(
      seats.separated == "yes" ~ TRUE,
      seats.separated == "no"  ~ FALSE,
      TRUE                     ~ NA
    ),
  
    
   # determining the state of a bench!
    is_inclusive = case_when(
      
    # --- exclusionary ---
    has_separated_seats == TRUE                                          ~ "hostile",
    has_armrest == TRUE                                                  ~ "hostile",
    lying_down == FALSE                                                  ~ "hostile",
    # tilted benches; no tag/attribute
    # sculptures in the middle, type of seat separator
  
    # --- non-hostile / sleep-friendly ---
    if(chooseBasemap == "municipality"){
      has_backrest == TRUE & has_armrest == FALSE & is_covered == TRUE     ~ "sleep-friendly_covered"},
    
    lying_down == TRUE                                                   ~ "sleep-friendly", # able to lie down; lack of is_covered is assumed to mean that it isn't covered
    has_backrest == TRUE & has_armrest == FALSE                          ~ "sleep-friendly", # inferred ability to lie down
    
    has_backrest == FALSE                          ~ "non-hostile", # no backrest, armrest and cover from rain
  
    TRUE                                                                 ~ NA_character_
  ))

cat("Inclusivity classification:\n")
table(inex_benches$is_inclusive, useNA = "always") # none have is_covered == True


```

## Visualisation incl. homeless shelters in Aarhus (ME)
OpenStreetMap does have a value for the amenity key called 'social_facility', which describes facilities providing social services such as group & nursing homes etc. in addition to someless shelters. And idk if we can extract something specific from a value (like social_facility), so we might have to find the data elsewhere or plot it ourselves (assuming that there would only be so mnay homeless shelters).

-> i assume we are thinking specifically of places where sleep is possible(?); otherwise there are also various 'væresteder' but i don't think they give beds for the night

_examples found of homeless shelters in Aarhus (DK: herberger (specifically the sleeping arrangements) / varmestue ("a place to go") / hjemløsetilbud)_
- Aarhus Vinterherberg (specifically for non-Danish homeless people AND only in the winter months) (blaakors.dk/vores-arbejde/noedherberger)
- Varmestue Herning (in Herning, so only important if examining Aarhus Commune)
- Spor 7 (Godsbanen; natvarmestue)
- Kirkens Korshær Herberget Tre Ege
- Omsorgscenter Pitstop (viby; omsorgscenter for  syge hjemløse og funktionelt hjemløse i Aarhus Kommune) (rodekors.dk/vores-arbejde/socialt-arbejde/hjemloese/omsorgscenter-pitstop?gad_source=1&gad_campaignid=23814314290&gclid=CjwKCAjw5NvPBhAoEiwA_2egfiD_SeISgOoQMtUX1mCNiw08K9-zdqpzBCMfNZgBMhWqXyEHQQ19wxoCEQ4QAvD_BwE)


_potentially relevant sources_
- information about 'Varmestuer' in DK by Kirkens Korshær: https://kirkenskorshaer.dk/2025/09/varmestuer/
- 'når menneskesynet vakler: Aarhus svigter de mest udsatte': https://jyllands-posten.dk/jpaarhus/debat/breve/ECE18602465/naar-menneskesynet-vakler-aarhus-svigter-de-mest-udsatte/


```{r visual_assessment}

# prepare bench data
benches_hos <- filter(inex_benches, is_inclusive == 'hostile')
benches_nonHos <- filter(inex_benches, is_inclusive == 'non-hostile')
benches_friend <- filter(inex_benches, is_inclusive == 'sleep-friendly')
benches_unknown   <- filter(inex_benches, is.na(is_inclusive))

# plot without NA benches
mapview(baseMap, layer.name = "Base Map") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")

# plot with NA benches
mapview(baseMap, layer.name = "Base Map") + mapview(benches_unknown, cex = 3, col.regions = "ivory2", alpha = 0.1, layer.name = "NA") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")



# plot with a few homeless shelters offering sleeping arrangements
vinterherberg_aarhus <-sf::st_sfc(sf::st_point(c(10.188392854268452,56.201760608688036)), crs = 4326)
spor7_aarhus <-sf::st_sfc(sf::st_point(c(10.18941672755415,56.151407205285174)), crs = 4326)
treEge_aarhus <-sf::st_sfc(sf::st_point(c(10.086763310082667,56.1546412110082)), crs = 4326)
omsorgscenter_aarhus <-sf::st_sfc(sf::st_point(c(10.158931423570165,56.11486169632215)), crs = 4326)


mapview(baseMap, layer.name = "Base Map") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly") + mapview(vinterherberg_aarhus, col.regions = "blue", cex = 5, layer.name = "HomelessShelter_vinterherberg") + mapview(spor7_aarhus, col.regions = "blue", cex = 5, layer.name = "HomelessShelter_spor7") + mapview(treEge_aarhus, col.regions = "blue", cex = 5, layer.name = "HomelessShelter_treEge") + mapview(omsorgscenter_aarhus, col.regions = "blue", cex = 5, layer.name = "HomelessShelter_omsorgscenter")


```


3) Noise Pollution Data

_DO NOT RUN THIS_
```{r prepare_noise_pollution_data}

# load the shapefile for noise pollution from streets during the night at 1.5 meters height
raw_streetNoise <- st_read("data/noisePollution/dk_2022_vej_nat_1_5m.shp") # downloaded from Miljøstyrelsen at: https://mst.dk/erhverv/tilskud-miljoeviden-og-data/data-og-databaser/miljoegis-data-om-natur-og-miljoe-paa-webkort/hent-data-udstillet-paa-miljoegis


# convert CRS to WGS84 (4326)
raw_streetNoise <- st_transform(raw_streetNoise, baseMap_crs)

print(st_crs( raw_streetNoise ))


raw_streetNoise <- st_make_valid(raw_streetNoise)
baseMap <- st_make_valid(baseMap)


# only keep data within Aarhus municipality
aarhus_night_streetNoise <- st_intersection(raw_streetNoise, baseMap)



# save the data as a GeoPackage; help from: https://mapping-in-r-workshop.ryanpeek.org/02_import_export_gpkg
bench_file <- sprintf("data/aarhus_night_streetNoise_%s.gpkg", chooseBasemap)
st_write(aarhus_night_streetNoise, bench_file, delete_dsn = TRUE)

```

_RUN THIS_
```{r load_prepared_noise_pollution_data}

# read in the noise pollution data for Aarhus only
aarhus_night_streetNoise <- st_read(bench_file)



```



```{r visualise_noisePollution_inAarhus}

aarhus_night_streetNoise <- aarhus_night_streetNoise %>% mutate( noise_band = paste0(isov1, "-", isov2, " dB") ) # creating a column with labels for the plot with the dB intervals

# static plot
ggplot(aarhus_night_streetNoise) +
  geom_sf(aes(fill = factor(isov1))) +
  labs(fill = "Noise level (dB)")


# interactive plot
mapview(
  aarhus_night_streetNoise,
  zcol = "noise_band")


```





```{r streetNoise_counts}

# areas with 50 or above db from street noise
aarhus_night_streetNoise_50 <- filter(aarhus_night_streetNoise, isov1>=50) # areas (i.e., rows) with decibel at 50 or above

bench_aarhus_night_streetNoise_50 <- st_intersection(inex_benches, aarhus_night_streetNoise_50) # intersection to get the benches with said areas


benches_inclusivity_streetNoise_50_count <- bench_aarhus_night_streetNoise_50 %>% st_drop_geometry() %>% count(is_inclusive) # count number of rows per is_inclusive category of these benches subjected to 50 or above db
benches_inclusivity_streetNoise_50_count <- rename(benches_inclusivity_streetNoise_50_count, count_50 = n)

head(benches_inclusivity_streetNoise_50_count)


# areas with below 50 db from street noise
aarhus_night_streetNoise_below50 <- filter(aarhus_night_streetNoise, isov1<50) # areas (i.e., rows) with decibel below 50

bench_aarhus_night_streetNoise_below50 <- st_intersection(inex_benches, aarhus_night_streetNoise_below50) # intersection to get the benches with said areas


benches_inclusivity_streetNoise_below50_count <- bench_aarhus_night_streetNoise_below50 %>% st_drop_geometry() %>% count(is_inclusive)
benches_inclusivity_streetNoise_below50_count <- rename(benches_inclusivity_streetNoise_below50_count, count_below50 = n)


head(benches_inclusivity_streetNoise_below50_count)
```

```{r streetNoise_proportions}
# proportion of benches pr. bench type (hostile, non-hostile, sleep-friendly) subjected to 50 or above db
benches_inclusivity_streetNoise <- merge(benches_inclusivity_streetNoise_50_count, benches_inclusivity_streetNoise_below50_count, by = "is_inclusive") # merge


total_counts <- inex_benches %>%  st_drop_geometry() %>% count(is_inclusive, name = "totalBenchCount")

benches_inclusivity_streetNoise <- merge(benches_inclusivity_streetNoise, total_counts, by = "is_inclusive")

benches_inclusivity_streetNoise <- mutate(benches_inclusivity_streetNoise,
                                          proportion50andAbove = count_50/totalBenchCount)

head(benches_inclusivity_streetNoise)


```

-> consider benches falling between / intersecting multiple polygons (noise)



### Reclassify benches as hostile, non-hostile and sleep-friendly (AR's classification code, ME's reclassification use)
```{r reclassification_of_benches}



for (bench in 1:nrow(inex_benches)){
  if (inex_benches$osm_id[bench] %in% bench_aarhus_night_streetNoise_50$osm_id & is.na(inex_benches$is_inclusive[bench])==FALSE){
    inex_benches$is_inclusive[bench] = 'hostile'
  }
}

cat("Inclusivity classification w. noise pollution:\n")
table(inex_benches$is_inclusive, useNA = "always")


```

```{r visual_assessment2}

# prepare bench data
benches_hos <- filter(inex_benches, is_inclusive == 'hostile')
benches_nonHos <- filter(inex_benches, is_inclusive == 'non-hostile')
benches_friend <- filter(inex_benches, is_inclusive == 'sleep-friendly')
benches_friendCover <- filter(inex_benches, is_inclusive == 'sleep-friendly_covered')
benches_unknown   <- filter(inex_benches, is.na(is_inclusive))

# plot without NA benches
mapview(baseMap, layer.name = "Base Map") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")

# plot with NA benches
mapview(baseMap, layer.name = "Base Map") + mapview(benches_unknown, cex = 3, col.regions = "ivory2", alpha = 0.1, layer.name = "NA") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")



# plot with a few homeless shelters offering sleeping arrangements
vinterherberg_aarhus <-sf::st_sfc(sf::st_point(c(10.188392854268452,56.201760608688036)), crs = 4326)
spor7_aarhus <-sf::st_sfc(sf::st_point(c(10.18941672755415,56.151407205285174)), crs = 4326)
treEge_aarhus <-sf::st_sfc(sf::st_point(c(10.086763310082667,56.1546412110082)), crs = 4326)
omsorgscenter_aarhus <-sf::st_sfc(sf::st_point(c(10.158931423570165,56.11486169632215)), crs = 4326)


mapview(baseMap, layer.name = "Base Map") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly") + mapview(vinterherberg_aarhus, col.regions = "blue", cex = 5, layer.name = "HomelessShelter_vinterherberg") + mapview(spor7_aarhus, col.regions = "blue", cex = 5, layer.name = "HomelessShelter_spor7") + mapview(treEge_aarhus, col.regions = "blue", cex = 5, layer.name = "HomelessShelter_treEge") + mapview(omsorgscenter_aarhus, col.regions = "blue", cex = 5, layer.name = "HomelessShelter_omsorgscenter")


```




 - - - - -  - - - - -

# Same Analysis as Above but for Aarhus Ringvej
```{r}

# load the shapefile for noise pollution from streets during the night at 1.5 meters height
aarhus_ring <- read_sf("data/ringVejGade/ringVejGade.geojson")

aarhus_ringvej <- filter(aarhus_ring, Ringvejen == 1)
aarhus_ringgade <- filter(aarhus_ring, Ringgaden == 1)


# intersect with aarhus municipality base map
aarhus_ringvej <- st_intersection(aarhus_ringvej, baseMap)
aarhus_ringgade <- st_intersection(aarhus_ringgade, baseMap)

mapview(aarhus_ringvej)

```

```{r download_benches_data_ring}

## inspiration from class scripts (specifically, class 06, which is based on luukvdmeer.github.io/UseR2024/notebook.html)


# query the roads within our area (i.e., Aarhus commune) from OpenStreetMap
assign("has_internet_via_proxy", TRUE, environment(curl::has_internet)) # telling R that we do in fact have wifi


# query the benches within our area (i.e., Aarhus commune) from OpenStreetMap
benches_ring <- osmdata::opq(sf::st_bbox(aarhus_ringvej)) %>% # generate bounding box
  osmdata::add_osm_feature(key = "amenity", value = "bench") %>%
  osmdata::osmdata_sf() %>%
  purrr::pluck("osm_points") %>%
  sf::st_intersection(aarhus_ringvej) # fitting to Aarhus municipality


# save the road data
saveRDS(benches_ring, "data/benches_osm_ringvej.rds")

```

```{r load_in_the_benches_data_if_already_computed_ring}

benches_ring <- readRDS("data/benches_osm_ringvej.rds")

mapview(aarhus_ringvej) + mapview(benches_ring)

```



```{r removing_columns_ring}


benches_ring <- subset(benches_ring, select = -c(artist_name, artwork_type, bin, colour, direction, material, note, panoramax, survey.date, tourism))


```

```{r classify_benches_ring} 

## CREATED BY AR

# now we create some basis for classifying the benches into hostile, non-hostile and sleep-friendly groups

inex_benches_ring <- benches_ring %>% 
  dplyr::mutate(
  
    # backrest
    has_backrest = case_when(
      backrest == "yes" ~ TRUE,
      backrest == "no"  ~ FALSE,
      TRUE              ~ NA   # anything else (including NA) stays unknown
    ),
    
    # armrest
    has_armrest = case_when(
      armrest == "yes" ~ TRUE,
      armrest == "no"  ~ FALSE,
      TRUE             ~ NA
    ),
    
    # can lie down
    can_lie_down = case_when(
      lying_down == "yes" ~ TRUE,
      lying_down == "no"  ~ FALSE,
      TRUE                ~ NA
    ),

    
    # separated seats (armrests between seats — exclusionary for rough sleeping)
    has_separated_seats = case_when(
      seats.separated == "yes" ~ TRUE,
      seats.separated == "no"  ~ FALSE,
      TRUE                     ~ NA
    ),
  
    
   # determining the state of a bench!
    is_inclusive = case_when(
      
    # --- exclusionary ---
    has_separated_seats == TRUE                                          ~ "hostile",
    has_armrest == TRUE                                                  ~ "hostile",
    lying_down == FALSE                                                  ~ "hostile",
    # tilted benches; no tag/attribute
    # sculptures in the middle, type of seat separator
  
    # --- non-hostile / sleep-friendly ---
    lying_down == TRUE                                                   ~ "sleep-friendly", # able to lie down; lack of is_covered is assumed to mean that it isn't covered
    has_backrest == TRUE & has_armrest == FALSE                          ~ "sleep-friendly", # inferred ability to lie down
    
    has_backrest == FALSE                          ~ "non-hostile", # no backrest, armrest and cover from rain
  
    TRUE                                                                 ~ NA_character_
  ))

cat("Inclusivity classification:\n")
table(inex_benches_ring$is_inclusive, useNA = "always")




```



```{r visual_assessment_ring}

# prepare bench data
benches_hos_ring <- filter(inex_benches_ring, is_inclusive == 'hostile')
benches_nonHos_ring <- filter(inex_benches_ring, is_inclusive == 'non-hostile')
benches_friend_ring <- filter(inex_benches_ring, is_inclusive == 'sleep-friendly')
benches_friendCover_ring <- filter(inex_benches_ring, is_inclusive == 'sleep-friendly_covered')
benches_unknown_ring   <- filter(inex_benches_ring, is.na(is_inclusive))

# plot without NA benches
mapview(aarhus_ringvej, layer.name = "Base Map") + mapview(benches_hos_ring, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos_ring, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend_ring, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")

# plot with NA benches
mapview(aarhus_ringvej, layer.name = "Base Map") + mapview(benches_unknown_ring, cex = 3, col.regions = "ivory2", alpha = 0.1, layer.name = "NA") + mapview(benches_hos_ring, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos_ring, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend_ring, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")


```



```{r noise_pollution_ring}
# only keep data within Aarhus municipality, ringvej
aarhus_night_streetNoise_ring <- st_intersection(raw_streetNoise, aarhus_ringvej)

# save the data as a GeoPackage; help from: https://mapping-in-r-workshop.ryanpeek.org/02_import_export_gpkg
st_write(aarhus_night_streetNoise_ring, "data/aarhus_night_streetNoise_ringvej.rds.gpkg", delete_dsn = TRUE)


```

```{r noise_pollution_readData_ring}

# read in the noise pollution data for Aarhus only
aarhus_night_streetNoise_ring <- st_read("data/aarhus_night_streetNoise_ringvej.rds.gpkg")


```


```{r streetNoise_counts_ring}

# areas with 50 or above db from street noise
aarhus_night_streetNoise_50_ring <- filter(aarhus_night_streetNoise_ring, isov1>=50) # areas (i.e., rows) with decibel at 50 or above

bench_aarhus_night_streetNoise_50_ring <- st_intersection(inex_benches_ring, aarhus_night_streetNoise_50_ring) # intersection to get the benches with said areas


benches_inclusivity_streetNoise_50_count_ring <- bench_aarhus_night_streetNoise_50_ring %>% st_drop_geometry() %>% count(is_inclusive) # count number of rows per is_inclusive category of these benches subjected to 50 or above db
benches_inclusivity_streetNoise_50_count_ring <- rename(benches_inclusivity_streetNoise_50_count_ring, count_50 = n)

head(benches_inclusivity_streetNoise_50_count_ring)


# areas with below 50 db from street noise
aarhus_night_streetNoise_below50_ring <- filter(aarhus_night_streetNoise_ring, isov1<50) # areas (i.e., rows) with decibel below 50

bench_aarhus_night_streetNoise_below50_ring <- st_intersection(inex_benches_ring, aarhus_night_streetNoise_below50_ring) # intersection to get the benches with said areas


benches_inclusivity_streetNoise_below50_count_ring <- bench_aarhus_night_streetNoise_below50_ring %>% st_drop_geometry() %>% count(is_inclusive)
benches_inclusivity_streetNoise_below50_count_ring <- rename(benches_inclusivity_streetNoise_below50_count_ring, count_below50 = n)


head(benches_inclusivity_streetNoise_below50_count_ring)
```

```{r streetNoise_proportions_ring}
# proportion of benches pr. bench type (hostile, non-hostile, sleep-friendly) subjected to 50 or above db
benches_inclusivity_streetNoise_ring <- merge(benches_inclusivity_streetNoise_50_count_ring, benches_inclusivity_streetNoise_below50_count_ring, by = "is_inclusive") # merge


total_counts_ring <- inex_benches_ring %>%  st_drop_geometry() %>% count(is_inclusive, name = "totalBenchCount")

benches_inclusivity_streetNoise_ring <- merge(benches_inclusivity_streetNoise_ring, total_counts_ring, by = "is_inclusive")

benches_inclusivity_streetNoise_ring <- mutate(benches_inclusivity_streetNoise_ring,
                                          proportion50andAbove = count_50/totalBenchCount)

head(benches_inclusivity_streetNoise_ring)


```



### Reclassify benches as hostile, non-hostile and sleep-friendly (AR's classification code, ME's reclassification use)
```{r reclassification_of_benches_ring}



for (bench in 1:nrow(inex_benches_ring)){
  if (inex_benches_ring$osm_id[bench] %in% bench_aarhus_night_streetNoise_50_ring$osm_id & is.na(inex_benches_ring$is_inclusive[bench])==FALSE){
    inex_benches_ring$is_inclusive[bench] = 'hostile'
  }
}

cat("Inclusivity classification w. noise pollution:\n")
table(inex_benches_ring$is_inclusive, useNA = "always")


```
```{r}
benches_hos_ring <- filter(inex_benches_ring, is_inclusive == 'hostile')
benches_nonHos_ring <- filter(inex_benches_ring, is_inclusive == 'non-hostile')
benches_friend_ring <- filter(inex_benches_ring, is_inclusive == 'sleep-friendly')
benches_friendCover_ring <- filter(inex_benches_ring, is_inclusive == 'sleep-friendly_covered')
benches_unknown_ring   <- filter(inex_benches_ring, is.na(is_inclusive))

# plot without NA benches
mapview(aarhus_ringvej, layer.name = "Base Map") + mapview(benches_hos_ring, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos_ring, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend_ring, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")


```

```{r}

benches_hos_ring <- filter(inex_benches_ring, is_inclusive == 'hostile')
benches_nonHos_ring <- filter(inex_benches_ring, is_inclusive == 'non-hostile')
benches_friend_ring <- filter(inex_benches_ring, is_inclusive == 'sleep-friendly')
benches_friendCover_ring <- filter(inex_benches_ring, is_inclusive == 'sleep-friendly_covered')
benches_unknown_ring   <- filter(inex_benches_ring, is.na(is_inclusive))

# plot with NA benches
mapview(aarhus_ringvej, layer.name = "Base Map") + mapview(benches_unknown_ring, cex = 3, col.regions = "ivory2", alpha = 0.1, layer.name = "NA") + mapview(benches_hos_ring, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos_ring, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend_ring, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")

```


```{r visual_assessment_ring2}

# prepare bench data
benches_hos_ring <- filter(inex_benches_ring, is_inclusive == 'hostile')
benches_nonHos_ring <- filter(inex_benches_ring, is_inclusive == 'non-hostile')
benches_friend_ring <- filter(inex_benches_ring, is_inclusive == 'sleep-friendly')
benches_friendCover_ring <- filter(inex_benches_ring, is_inclusive == 'sleep-friendly_covered')
benches_unknown_ring   <- filter(inex_benches_ring, is.na(is_inclusive))

# plot without NA benches
mapview(aarhus_ringvej, layer.name = "Base Map") + mapview(benches_hos_ring, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos_ring, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend_ring, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")

# plot with NA benches
mapview(aarhus_ringvej, layer.name = "Base Map") + mapview(benches_unknown_ring, cex = 3, col.regions = "ivory2", alpha = 0.1, layer.name = "NA") + mapview(benches_hos_ring, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos_ring, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend_ring, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")


```






 - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - - 


## Voronoi - Service Areas (zoom to study area(s)) (AR)

Need to adjust bbox coordinates for our chosen area - this is incredibly large and uninformative right now.
Can use mapview(baseMap_utm) to find the (approximately) right values interactively.

Essentially, this section is just kind of a proof-of-concept right now!

```{r, voronoi study area}

inex_benches <- st_transform(inex_benches, 25832) # transform crs to projected crs with meter units; here EPSG: 25832 (UTM 32N), which is used for Europe; before the WGS84 (4326) CRS was used
baseMap_utm <- st_transform( baseMap, crs=25832 )


# prepare bench data with new crs
benches_hos <- filter(inex_benches, is_inclusive == 'hostile')
benches_nonHos <- filter(inex_benches, is_inclusive == 'non-hostile')
benches_friend <- filter(inex_benches, is_inclusive == 'sleep-friendly')
benches_friendCover <- filter(inex_benches, is_inclusive == 'sleep-friendly_covered')
benches_unknown   <- filter(inex_benches, is.na(is_inclusive))


study_bbox <- st_bbox(
  c(xmin = 566000, ymin = 6220000, xmax = 572000, ymax = 6226000),  # <- adjust
  crs = st_crs(baseMap_utm)
) %>% st_as_sfc()

benches_study_all <- st_intersection(inex_benches, study_bbox)
benches_study_hos  <- st_intersection(benches_hos, study_bbox)
benches_study_nonHos  <- st_intersection(benches_nonHos, study_bbox)
benches_study_friend  <- st_intersection(benches_friend, study_bbox)
benches_study_friendCover  <- st_intersection(benches_friendCover, study_bbox)
baseMap_study     <- st_intersection(baseMap_utm, study_bbox)

cat(sprintf("Benches in study area: %d total | %d hostile | %d non-hostile | %d sleep-friendly | %d sleep-friendly_covered | %d NA\n",
  nrow(benches_study_all),
  nrow(benches_study_hos),
  nrow(benches_study_nonHos),
  nrow(benches_study_friend),
  nrow(benches_study_friendCover),
  nrow(benches_study_all) - nrow(benches_study_hos) - nrow(benches_study_nonHos) - nrow(benches_study_friend) - nrow(benches_study_friendCover)
))



```


```{r voronoi-polygons1}

# keep only benches with a definite inclusive/exclusionary label
benches_marked <- dplyr::filter(inex_benches, !is.na(is_inclusive))

# hostile, non-hostile, friendly, friendlycovered
voronoi_raw <- benches_marked %>%
  st_geometry() %>%
  st_union() %>%
  st_voronoi() %>%
  st_collection_extract("POLYGON") %>%
  st_sf() %>%
  st_intersection(st_make_valid(baseMap_utm))

voronoi_labelled <- voronoi_raw %>%
  st_join(benches_marked["is_inclusive"], join = st_nearest_feature) %>%
  mutate(
    area_m2  = as.numeric(st_area(geometry)),
    category = ifelse(is_inclusive == "hostile", "Hostile", ifelse(is_inclusive == "non-hostile", "Non-Hostile", ifelse(is_inclusive == "sleep-friendly", "Sleep-Friendly", "Sleep-Friendly_covered")))
  )




```

```{r voronoi-plot1, fig.width=9, fig.height=7}

voronoi_labelled <- voronoi_raw %>%
  st_join(benches_marked["is_inclusive"], join = st_nearest_feature) %>%
  mutate(
    area_m2  = as.numeric(st_area(geometry)),
    category = ifelse(is_inclusive == "hostile", "Hostile", ifelse(is_inclusive == "non-hostile", "Non-Hostile", ifelse(is_inclusive == "sleep-friendly", "Sleep-Friendly", "Sleep-Friendly_covered")))
  )


theme_ppt <- theme_void(base_family = "sans") +
  theme(
    plot.title      = element_text(face = "bold", size = 15, hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle   = element_text(colour = "grey40", size = 10, hjust = 0.5, margin = margin(b = 10)),
    legend.position = "bottom",
    legend.title    = element_text(size = 9),
    legend.text     = element_text(size = 8)
  )

ggplot() +
  geom_sf(data = voronoi_labelled, aes(fill = category),
          colour = "white", linewidth = 0.2, alpha = 0.7) +
  geom_sf(data = baseMap_utm, fill = NA, colour = "grey40", linewidth = 0.5) +
  geom_sf(data = benches_marked, aes(colour = is_inclusive), size = 1.2, show.legend = FALSE) +
  scale_fill_manual(values = c("Hostile" = "#ffc05c", "Non-Hostile" = "#FFF159", "Sleep-Friendly" = "lightgreen"),
                    name = "Service area") +
  scale_colour_manual(values = c("hostile" = "#C7701A", "non-hostile" = "yellow3", "sleep-friendly" = "green4")) +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5,
                             override.aes = list(alpha = 0.85, size = 5))) +
  labs(
    title    = "Who Does Each Bench Serve? (Aarhus Municipality, Denmark)",
    subtitle = "Voronoi Diagram - Each region shows the area served by its nearest bench"
  ) +
  theme_ppt


mapview(voronoi_labelled, zcol = "category")




```

```{r voronoi-summary1}

voronoi_labelled %>%
  st_drop_geometry() %>%
  group_by(category) %>%
  summarise(
    n         = n(),
    mean_m2   = round(mean(area_m2)),
    median_m2 = round(median(area_m2)),
    .groups = "drop"
  ) %>%
  print()

```



```{r voronoi-polygons2}

# Use ALL benches including NA
voronoi_raw <- inex_benches %>%
  st_geometry() %>%
  st_union() %>%
  st_voronoi() %>%
  st_collection_extract("POLYGON") %>%
  st_sf() %>%
  st_intersection(st_make_valid(baseMap_utm))

voronoi_labelled <- voronoi_raw %>%
  st_join(inex_benches["is_inclusive"], join = st_nearest_feature) %>%
  mutate(
    area_m2  = as.numeric(st_area(geometry)),
    category = case_when(
      is_inclusive == "hostile" ~ "Hostile",
      is_inclusive == "non-hostile" ~ "Non-Hostile",
      is_inclusive == "sleep-friendly" ~ "Sleep-Friendly",
      is_inclusive == "sleep-friendly_covered" ~ "Sleep-Friendly_covered",
      TRUE ~ "Unknown"
    )
  )

```


```{r voronoi-plot2, fig.width=9, fig.height=7}


theme_ppt <- theme_void(base_family = "sans") +
  theme(
    plot.title      = element_text(face = "bold", size = 15, hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle   = element_text(colour = "grey40", size = 10, hjust = 0.5, margin = margin(b = 10)),
    legend.position = "bottom",
    legend.title    = element_text(size = 9),
    legend.text     = element_text(size = 8)
  )

ggplot() +
  geom_sf(data = voronoi_labelled, aes(fill = category),
          colour = "white", linewidth = 0.2, alpha = 0.7) +
  geom_sf(data = baseMap_utm, fill = NA, colour = "grey40", linewidth = 0.5) +
  geom_sf(data = benches_marked, aes(colour = is_inclusive), size = 1.2, show.legend = FALSE) +
  scale_fill_manual(values = c("Hostile" = "#ffc05c", "Non-Hostile" = "#FFF159", "Sleep-Friendly" = "lightgreen", "Unknown" = "grey70"),
                    name = "Service area") +
  scale_colour_manual(values = c("hostile" = "#C7701A", "non-hostile" = "yellow3", "sleep-friendly" = "green4"),
                      na.value = "grey60") +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5,
                             override.aes = list(alpha = 0.85, size = 5))) +
  labs(
    title    = "Who Does Each Bench Serve? (Aarhus Municipality, Denmark)",
    subtitle = "Voronoi Diagram - Each region shows the area served by its nearest bench"
  ) +
  theme_ppt


mapview(voronoi_labelled, zcol = "category")




```


```{r voronoi-summary2}

voronoi_labelled %>%
  st_drop_geometry() %>%
  group_by(category) %>%
  summarise(
    n         = n(),
    mean_m2   = round(mean(area_m2)),
    median_m2 = round(median(area_m2)),
    .groups = "drop"
  ) %>%
  print()

```



```{r voronoi-interactive}

mapview(voronoi_labelled, zcol = "category",
        col.regions = c("Hostile" = "#ffc05c", "Non-Hostile" = "#FFF159", "Sleep-Friendly" = "lightgreen", "Unknown" = "grey70"),
        layer.name = "Service area") +
  
  mapview(inex_benches, zcol = "is_inclusive",
        col.regions = c("hostile" = "#C7701A", "non-hostile" = "yellow3", "sleep-friendly" = "green4"),
        na.color = "grey60",
        cex = 3, layer.name = "Benches")



```


# IDK anymore


```{r, voronoi summary}

# IDK what this is trying to do

#cat("=== All benches — service area (m²) ===\n")
#summary(voronoi_all$area_m2)

#cat("\n=== Inclusive benches only — service area (m²) ===\n")
#summary(voronoi_in$area_m2)

#ratio <- mean(voronoi_in$area_m2) / mean(voronoi_all$area_m2)
#cat(sprintf("\nMean service area is %.1fx larger when restricted to inclusive benches.\n", ratio))

```


## Demographic Overlay (AR) -> look into this

An idea for using dkstat data which is a part of the opengov dk intitative: https://github.com/rOpenGov/dkstat 
We could use IDs to identify different demographics spread through aarhus - not sure if an ID for e.g., 'homeless' or socialt udsatte exists or one for elderly people exists but it might be a nice (maybe) way to get an understanding of how the presence and distribution of benches say something about how we care (or not) about various demographics of people.

```{r, elderly}

#install.packages(
#  "dkstat",
#  repos = c(
#    ropengov = "https://ropengov.r-universe.dev",
#    getOption("repos")
#  )
#)

#meta <- dkstat::dst_meta("", lang = "en") # need to test an ID
#meta$variables

```




## HELP for Paper stuff

```{r}

pacman::p_load(attachment)
attachment::att_from_rmd("benchesBASIC.Rmd")

```

```{r, citation}

citation(package="osmdata")

```

