---
title: "MAIN"
course: "Cultural Datascience: Spatial Analytics at AU"
authors: "Aiswarya Roy & Mie Norre Engemann"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
---

_Script Description:_  
The current script (1) creates basemaps based one Aarhus municipality and Aarhus city (center) based on Aarhus ringvej and Aarhus ringgadde. Next, the script (2) downloads bench data from OpenStreetMap for the chosen basemap and classifies these benches as either hostile, non-hostile or sleep-friendly based on the provided tags from OpenStreetMap. Next, the script (3) uses road noise pollution data from city roads to re-classify the benches as hostile if they are subjected to at least 50 dB Lnight. Lastly, (4) the script creates Voronoi diagrams.

- For more information on the libraries used and the data sources, refer to the paper's appendix.
- The noise pollution data files are too big for GitHub and as such, only the noise pollution data for each basemap is saved in the data/ folder.

NB! To run the code, know that you cannot run the chunk loading the noise pollution data from Miljøstyrelsens MiljøGIS, due to the reason noted above. AND that the variable *chooseBasemap* is what decides the basemap used for each run of the script.


_Script:_

```{r setup, include=FALSE}
## Global options
library(pacman)

pacman::p_load(rmdformats, knitr)

options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r}

knitr::clean_cache()
chooseBasemap <- "ringvej" # options: ringvej, ringgade, municipality

```


## Loading Libraries

```{r load_libraries, message = FALSE, warning=FALSE}

library(pacman)

pacman::p_load(sf,
               tidyverse,
               osmdata,
               ggplot2,
               geodata,
               dplyr,
               geopackage,
               mapview)


```


## Data Preparation (ME)

Loading Data, CRS conversion etc.

1) Base Map of Aarhus
```{r download_municipality_data}

## BASEMAP;   inspiration from class scripts
workingDirectory_path <- getwd()

# get municipality data
municipalityData_downloaded <- geodata::gadm(country="Denmark", level=2, path=file.path(workingDirectory_path, "data"), version="3.6") # data from the GADM database and save to data/gadm/

municipalityData_sf <- st_as_sf(municipalityData_downloaded)


```

```{r prepare_municipality_base_map}

## BASEMAP;   inspiration from class scripts

# preparing the data for this project
municipalityData_sf$NAME_2[31] <- "Aarhus" # from århus to aarhus

municipalityData_sf <- subset(municipalityData_sf, select = -c(GID_1, NL_NAME_1, GID_2, VARNAME_2, NL_NAME_2, CC_2)) # removing columns not used


# filter to only extract Aarhus
aarhus_municipality <- filter(municipalityData_sf, NAME_2 == "Aarhus")


# save CRS
baseMap_crs = st_crs(aarhus_municipality) # longlat, WGS84

# view
#mapview(aarhus_municipality)


```


1.1) Ringvejen / Ringgaden (other option for basemap)
```{r}

# load the shapefile for noise pollution from streets during the night at 1.5 meters height
aarhus_ring <- read_sf("data/ringVejGade/ringVejGade.geojson") # custom polygons from geojson.io powered by Mapbox

aarhus_ringvej <- filter(aarhus_ring, Ringvejen == 1)
aarhus_ringgade <- filter(aarhus_ring, Ringgaden == 1)

#mapview(aarhus_ringvej)
#mapview(aarhus_ringgade)

```



2) Bench Data
```{r currentMap}

# setting the basemap to the chooseBasemap value (ringvej, ringgade or municipality)

if (chooseBasemap == "ringvej"){
  baseMap <- aarhus_ringvej
  
  benchRemoveCol <- c("artist_name", "artwork_type", "bin", "colour", "direction", "material", "note", "panoramax", "survey.date", "tourism")
  
} else if (chooseBasemap == "ringgade"){
  baseMap <- aarhus_ringgade
  benchRemoveCol <- c("artist_name", "artwork_type", "colour", "direction", "material", "panoramax", "survey.date", "tourism")
  
} else if(chooseBasemap == "municipality") {
  baseMap <- aarhus_municipality
  benchRemoveCol <- c("artist_name", "artwork_type", "bin", "colour", "inscription", "not.inscription", "direction", "lit", "material", "not.tourism.artwork", "note", "panoramax", "source", "survey.date", "tourism", "level")
} else{
  
    stop("Input ringvej, ringgade or municipality for the chooseBasemap variable")
  }


```


```{r download_benches_data}

## inspiration from class scripts (specifically, class 06, which is based on luukvdmeer.github.io/UseR2024/notebook.html)

assign("has_internet_via_proxy", TRUE, environment(curl::has_internet)) # telling R that we do in fact have wifi

# query the benches within our area (i.e., within the basemap) from OpenStreetMap
benches_aarhus <- osmdata::opq(sf::st_bbox(baseMap)) %>% # generate initial bounding box
  osmdata::add_osm_feature(key = "amenity", value = "bench") %>% # define the bench data
  osmdata::osmdata_sf() %>%
  purrr::pluck("osm_points") %>%
  sf::st_intersection(baseMap) # fitting to the basemap (Aarhus municipality, ringvejen or ringgaden)


# save the road data
#########################################file.remove("data/benches_osm.rds")
bench_file <- sprintf("data/benches_osm_%s.rds", chooseBasemap)
saveRDS(benches_aarhus, bench_file)


```


```{r load_in_the_benches_data_if_already_computed}

benches <- readRDS(bench_file)

```


```{r removing_columns}

benches <- dplyr::select(benches, -all_of(benchRemoveCol)) # 'lit', 'material' and other features could be incorporated into the classifications of hostile, non-hostile and sleep-friendly benches but wasn't for the current project

# initial visualisation of the benches on the basemap
#mapview(baseMap) + mapview(benches)

# saving the map
#unclassifiedBench_plot <- mapview(baseMap) + mapview(benches)
#unclassifiedBench_filepath <- sprintf("out/unclassifiedBenches_osm_%s.png", chooseBasemap)

```


_NOTE:_ As OpenStreetMap also uses the WGS84 (4326) CRS, there is no need for CRS conversion.


```{r classify_benches} 

## CREATED BY AR

# now we create a basis for classifying the benches into hostile, non-hostile and sleep-friendly groups

inex_benches <- benches %>% 
  dplyr::mutate(
  
    # backrest
    has_backrest = case_when(
      backrest == "yes" ~ TRUE,
      backrest == "no"  ~ FALSE,
      TRUE              ~ NA   # anything else (including NA) stays unknown
    ),
    
    # armrest
    has_armrest = case_when(
      armrest == "yes" ~ TRUE,
      armrest == "no"  ~ FALSE,
      TRUE             ~ NA
    ),
    
    # can lie down
    can_lie_down = case_when(
      lying_down == "yes" ~ TRUE,
      lying_down == "no"  ~ FALSE,
      TRUE                ~ NA
    ),
    
    # covered (shelter from rain); few benches have this tag and only outside of ringvejen
    if(chooseBasemap == "municipality"){
      is_covered = case_when(
        covered == "yes" ~ TRUE,
        covered == "no"  ~ FALSE,
        TRUE             ~ NA)
    },
    
    # separated seats (armrests between seats — exclusionary for rough sleeping)
    has_separated_seats = case_when(
      seats.separated == "yes" ~ TRUE,
      seats.separated == "no"  ~ FALSE,
      TRUE                     ~ NA
    ),
  
    
   # determining the state of a bench!
    is_inclusive = case_when(
      
    # --- exclusionary ---
    has_separated_seats == TRUE                                          ~ "hostile",
    has_armrest == TRUE                                                  ~ "hostile",
    lying_down == FALSE                                                  ~ "hostile",
    # tilted benches; no tag/attribute
    # sculptures in the middle, type of seat separator; no tag/attribute
  
    # --- non-hostile / sleep-friendly ---
    lying_down == TRUE                                                   ~ "sleep-friendly", # able to lie down; lack of is_covered is assumed to mean that it isn't covered
    has_backrest == TRUE & has_armrest == FALSE                          ~ "sleep-friendly", # inferred ability to lie down
    
    has_backrest == FALSE                          ~ "non-hostile", # no backrest, armrest and cover from rain
  
    TRUE                                                                 ~ NA_character_
  ))

cat("Inclusivity classification:\n")
table(inex_benches$is_inclusive, useNA = "always") # none have is_covered == True


# save
name_file <- sprintf("out/initial_benchClassification_%s.csv", chooseBasemap)
write.table(inex_benches$is_inclusive, file = name_file,
            sep = "\t", row.names = F)


```

```{r visual_assessment}

# prepare bench data
benches_hos <- filter(inex_benches, is_inclusive == 'hostile')
benches_nonHos <- filter(inex_benches, is_inclusive == 'non-hostile')
benches_friend <- filter(inex_benches, is_inclusive == 'sleep-friendly')
benches_unknown   <- filter(inex_benches, is.na(is_inclusive))

# plot without NA benches
#mapview(baseMap, layer.name = "Base Map") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")

# plot with NA benches
#mapview(baseMap, layer.name = "Base Map") + mapview(benches_unknown, cex = 3, col.regions = "ivory2", alpha = 0.1, layer.name = "NA") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")


```


3) Noise Pollution Data

_DO NOT RUN THIS_
```{r prepare_noise_pollution_data}

# load the shapefile for noise pollution from streets during the night at 1.5 meters height
raw_streetNoise <- st_read("data/noisePollution/dk_2022_vej_nat_1_5m.shp") # downloaded from Miljøstyrelsen at: https://mst.dk/erhverv/tilskud-miljoeviden-og-data/data-og-databaser/miljoegis-data-om-natur-og-miljoe-paa-webkort/hent-data-udstillet-paa-miljoegis


# convert CRS to WGS84 (4326)
raw_streetNoise <- st_transform(raw_streetNoise, baseMap_crs)


raw_streetNoise <- st_make_valid(raw_streetNoise)
baseMap <- st_make_valid(baseMap)


# only keep data within the basemap (Aarhus municipality, ringvejen or ringgaden)
aarhus_night_streetNoise <- st_intersection(raw_streetNoise, baseMap)


# save the data as a GeoPackage; help from: https://mapping-in-r-workshop.ryanpeek.org/02_import_export_gpkg
noise_file <- sprintf("data/aarhus_night_streetNoise_%s.gpkg", chooseBasemap)
st_write(aarhus_night_streetNoise, noise_file, delete_dsn = TRUE)

```

_RUN THIS_
```{r load_prepared_noise_pollution_data}

# read in the noise pollution data for Aarhus only
aarhus_night_streetNoise <- st_read(noise_file) # data from Miljøstyrelsen, downloaded form the MiljøGIS



```



```{r visualise_noisePollution_inAarhus}

aarhus_night_streetNoise <- aarhus_night_streetNoise %>% mutate( noise_band = paste0(isov1, "-", isov2, " dB") ) # creating a column with labels for the plot with the dB intervals

# static plot
ggplot(aarhus_night_streetNoise) +
  geom_sf(aes(fill = factor(isov1))) +
  labs(title="Noise Pollution from Road Noise within the Basemap", fill = "Noise level (dB)")


# interactive plot
#mapview( aarhus_night_streetNoise,  zcol = "noise_band")


```



Descriptive statistics for the benches subjected to road noise

```{r streetNoise_counts}

# areas with 50 or above db from street noise
aarhus_night_streetNoise_50 <- filter(aarhus_night_streetNoise, isov1>=50) # areas (i.e., rows) with decibel at 50 or above

bench_aarhus_night_streetNoise_50 <- st_intersection(inex_benches, aarhus_night_streetNoise_50) # intersection to get the benches with said areas


benches_inclusivity_streetNoise_50_count <- bench_aarhus_night_streetNoise_50 %>% st_drop_geometry() %>% count(is_inclusive) # count number of rows per is_inclusive category of these benches subjected to 50 or above db
benches_inclusivity_streetNoise_50_count <- rename(benches_inclusivity_streetNoise_50_count, count_50 = n)

head(benches_inclusivity_streetNoise_50_count)


# areas with below 50 db from street noise
aarhus_night_streetNoise_below50 <- filter(aarhus_night_streetNoise, isov1<50) # areas (i.e., rows) with decibel below 50

bench_aarhus_night_streetNoise_below50 <- st_intersection(inex_benches, aarhus_night_streetNoise_below50) # intersection to get the benches with said areas


benches_inclusivity_streetNoise_below50_count <- bench_aarhus_night_streetNoise_below50 %>% st_drop_geometry() %>% count(is_inclusive)
benches_inclusivity_streetNoise_below50_count <- rename(benches_inclusivity_streetNoise_below50_count, count_below50 = n)


head(benches_inclusivity_streetNoise_below50_count)

```

```{r streetNoise_proportions}

# proportion of benches pr. bench type (hostile, non-hostile, sleep-friendly) subjected to 50 or above db
benches_inclusivity_streetNoise <- merge(benches_inclusivity_streetNoise_50_count, benches_inclusivity_streetNoise_below50_count, by = "is_inclusive") # merge


total_counts <- inex_benches %>%  st_drop_geometry() %>% count(is_inclusive, name = "totalBenchCount")

benches_inclusivity_streetNoise <- merge(benches_inclusivity_streetNoise, total_counts, by = "is_inclusive")

benches_inclusivity_streetNoise <- mutate(benches_inclusivity_streetNoise,
                                          proportion50andAbove = count_50/totalBenchCount)

head(benches_inclusivity_streetNoise)


# save
name_file <- sprintf("out/benchesVSnoisePollution_%s.csv", chooseBasemap)
write.table(benches_inclusivity_streetNoise, file = name_file,
            sep = "\t", row.names = F)

```

-> consider benches falling between / intersecting multiple polygons (noise)



### Reclassify benches as hostile, non-hostile and sleep-friendly (AR's classification code, ME's reclassification use)
```{r reclassification_of_benches}

for (bench in 1:nrow(inex_benches)){
  if (inex_benches$osm_id[bench] %in% bench_aarhus_night_streetNoise_50$osm_id & is.na(inex_benches$is_inclusive[bench])==FALSE){
    inex_benches$is_inclusive[bench] = 'hostile'
  }
}

cat("Inclusivity classification w. noise pollution:\n")
table(inex_benches$is_inclusive, useNA = "always")

# save
name_file <- sprintf("out/afterNoise_benchClassification_%s.csv", chooseBasemap)
write.table(inex_benches$is_inclusive, file = name_file,
            sep = "\t", row.names = F)


```

```{r visual_assessment2}

# prepare bench data
benches_hos <- filter(inex_benches, is_inclusive == 'hostile')
benches_nonHos <- filter(inex_benches, is_inclusive == 'non-hostile')
benches_friend <- filter(inex_benches, is_inclusive == 'sleep-friendly')
benches_friendCover <- filter(inex_benches, is_inclusive == 'sleep-friendly_covered')
benches_unknown   <- filter(inex_benches, is.na(is_inclusive))

# plot without NA benches
#mapview(baseMap, layer.name = "Base Map") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")

# plot with NA benches
#mapview(baseMap, layer.name = "Base Map") + mapview(benches_unknown, cex = 3, col.regions = "ivory2", alpha = 0.1, layer.name = "NA") + mapview(benches_hos, cex = 3, col.regions = "orange", alpha = 0.1, layer.name = "Hostile") + mapview(benches_nonHos, cex = 3, col.regions = "#FFF159", alpha = 0.1, layer.name = "Non-Hostile")  + mapview(benches_friend, cex = 3, col.regions = "lightgreen", alpha = 0.1, layer.name = "Sleep-Friendly")


```




 - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - -  - - - - - 


## 4) Voronoi - Service Areas (zoom to study area(s)) (AR)

Essentially, this section is just kind of a proof-of-concept right now!

```{r, voronoi study area}

# transform crs to projected crs with meter units; here EPSG: 25832 (UTM 32N), which is used for Europe; before the WGS84 (4326) CRS was used
inex_benches <- st_transform(inex_benches, 25832)
baseMap_utm <- st_transform( baseMap, crs=25832 )


# prepare bench data with new crs
benches_hos <- filter(inex_benches, is_inclusive == 'hostile')
benches_nonHos <- filter(inex_benches, is_inclusive == 'non-hostile')
benches_friend <- filter(inex_benches, is_inclusive == 'sleep-friendly')
benches_friendCover <- filter(inex_benches, is_inclusive == 'sleep-friendly_covered')
benches_unknown   <- filter(inex_benches, is.na(is_inclusive))


study_bbox <- st_bbox(
  c(xmin = 566000, ymin = 6220000, xmax = 572000, ymax = 6226000),  # <- adjust
  crs = st_crs(baseMap_utm)
) %>% st_as_sfc()

benches_study_all <- st_intersection(inex_benches, study_bbox)
benches_study_hos  <- st_intersection(benches_hos, study_bbox)
benches_study_nonHos  <- st_intersection(benches_nonHos, study_bbox)
benches_study_friend  <- st_intersection(benches_friend, study_bbox)
benches_study_friendCover  <- st_intersection(benches_friendCover, study_bbox)
baseMap_study     <- st_intersection(baseMap_utm, study_bbox)

cat(sprintf("Benches in study area: %d total | %d hostile | %d non-hostile | %d sleep-friendly | %d sleep-friendly_covered | %d NA\n",
  nrow(benches_study_all),
  nrow(benches_study_hos),
  nrow(benches_study_nonHos),
  nrow(benches_study_friend),
  nrow(benches_study_friendCover),
  nrow(benches_study_all) - nrow(benches_study_hos) - nrow(benches_study_nonHos) - nrow(benches_study_friend) - nrow(benches_study_friendCover)
))



```


```{r voronoi-polygons1}

# keep only benches with a definite inclusive/exclusionary label
benches_marked <- dplyr::filter(inex_benches, !is.na(is_inclusive))

# hostile, non-hostile, friendly, friendly_covered
voronoi_raw <- benches_marked %>%
  st_geometry() %>%
  st_union() %>%
  st_voronoi() %>%
  st_collection_extract("POLYGON") %>%
  st_sf() %>%
  st_intersection(st_make_valid(baseMap_utm))

voronoi_labelled <- voronoi_raw %>%
  st_join(benches_marked["is_inclusive"], join = st_nearest_feature) %>%
  mutate(
    area_m2  = as.numeric(st_area(geometry)),
    category = ifelse(is_inclusive == "hostile", "Hostile", ifelse(is_inclusive == "non-hostile", "Non-Hostile", ifelse(is_inclusive == "sleep-friendly", "Sleep-Friendly", "Sleep-Friendly_covered")))
  )




```

```{r voronoi-plot1, fig.width=9, fig.height=7}

voronoi_labelled <- voronoi_raw %>%
  st_join(benches_marked["is_inclusive"], join = st_nearest_feature) %>%
  mutate(
    area_m2  = as.numeric(st_area(geometry)),
    category = ifelse(is_inclusive == "hostile", "Hostile", ifelse(is_inclusive == "non-hostile", "Non-Hostile", ifelse(is_inclusive == "sleep-friendly", "Sleep-Friendly", "Sleep-Friendly_covered")))
  )


theme_ppt <- theme_void(base_family = "sans") +
  theme(
    plot.title      = element_text(face = "bold", size = 15, hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle   = element_text(colour = "grey40", size = 10, hjust = 0.5, margin = margin(b = 10)),
    legend.position = "bottom",
    legend.title    = element_text(size = 9),
    legend.text     = element_text(size = 8)
  )

ggplot() +
  geom_sf(data = voronoi_labelled, aes(fill = category),
          colour = "white", linewidth = 0.2, alpha = 0.7) +
  geom_sf(data = baseMap_utm, fill = NA, colour = "grey40", linewidth = 0.5) +
  geom_sf(data = benches_marked, aes(colour = is_inclusive), size = 1.2, show.legend = FALSE) +
  scale_fill_manual(values = c("Hostile" = "#ffc05c", "Non-Hostile" = "#FFF159", "Sleep-Friendly" = "lightgreen"),
                    name = "Service area") +
  scale_colour_manual(values = c("hostile" = "#C7701A", "non-hostile" = "yellow3", "sleep-friendly" = "green4")) +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5,
                             override.aes = list(alpha = 0.85, size = 5))) +
  labs(
    title    = "Who Does Each Bench Serve? (Aarhus Municipality, Denmark)",
    subtitle = "Voronoi Diagram - Each region shows the area served by its nearest bench"
  ) +
  theme_ppt


mapview(voronoi_labelled, zcol = "category")


```

```{r voronoi-summary1}

voronoi_labelled %>%
  st_drop_geometry() %>%
  group_by(category) %>%
  summarise(
    n         = n(),
    mean_m2   = round(mean(area_m2)),
    median_m2 = round(median(area_m2)),
    .groups = "drop"
  ) %>%
  print()

```



```{r voronoi-polygons2}

# Use ALL benches including NA
voronoi_raw <- inex_benches %>%
  st_geometry() %>%
  st_union() %>%
  st_voronoi() %>%
  st_collection_extract("POLYGON") %>%
  st_sf() %>%
  st_intersection(st_make_valid(baseMap_utm))

voronoi_labelled <- voronoi_raw %>%
  st_join(inex_benches["is_inclusive"], join = st_nearest_feature) %>%
  mutate(
    area_m2  = as.numeric(st_area(geometry)),
    category = case_when(
      is_inclusive == "hostile" ~ "Hostile",
      is_inclusive == "non-hostile" ~ "Non-Hostile",
      is_inclusive == "sleep-friendly" ~ "Sleep-Friendly",
      is_inclusive == "sleep-friendly_covered" ~ "Sleep-Friendly_covered",
      TRUE ~ "Unknown"
    )
  )

```


```{r voronoi-plot2, fig.width=9, fig.height=7}


theme_ppt <- theme_void(base_family = "sans") +
  theme(
    plot.title      = element_text(face = "bold", size = 15, hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle   = element_text(colour = "grey40", size = 10, hjust = 0.5, margin = margin(b = 10)),
    legend.position = "bottom",
    legend.title    = element_text(size = 9),
    legend.text     = element_text(size = 8)
  )

ggplot() +
  geom_sf(data = voronoi_labelled, aes(fill = category),
          colour = "white", linewidth = 0.2, alpha = 0.7) +
  geom_sf(data = baseMap_utm, fill = NA, colour = "grey40", linewidth = 0.5) +
  geom_sf(data = benches_marked, aes(colour = is_inclusive), size = 1.2, show.legend = FALSE) +
  scale_fill_manual(values = c("Hostile" = "#ffc05c", "Non-Hostile" = "#FFF159", "Sleep-Friendly" = "lightgreen", "Unknown" = "grey70"),
                    name = "Service area") +
  scale_colour_manual(values = c("hostile" = "#C7701A", "non-hostile" = "yellow3", "sleep-friendly" = "green4"),
                      na.value = "grey60") +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5,
                             override.aes = list(alpha = 0.85, size = 5))) +
  labs(
    title    = "Who Does Each Bench Serve? (Aarhus Municipality, Denmark)",
    subtitle = "Voronoi Diagram - Each region shows the area served by its nearest bench"
  ) +
  theme_ppt


mapview(voronoi_labelled, zcol = "category")

```


```{r voronoi-summary2}

voronoi_labelled %>%
  st_drop_geometry() %>%
  group_by(category) %>%
  summarise(
    n         = n(),
    mean_m2   = round(mean(area_m2)),
    median_m2 = round(median(area_m2)),
    .groups = "drop"
  ) %>%
  print()

# save
voronoi_summary <- voronoi_labelled %>%
  st_drop_geometry() %>%
  group_by(category) %>%
  summarise(
    n         = n(),
    mean_m2   = round(mean(area_m2)),
    median_m2 = round(median(area_m2)),
    .groups = "drop"
  ) %>%
  print()

name_file <- sprintf("out/voronoi_summary_%s.csv", chooseBasemap)
write.table(voronoi_summary, file = name_file,
            sep = "\t", row.names = F)

```



```{r voronoi-interactive}

mapview(voronoi_labelled, zcol = "category",
        col.regions = c("Hostile" = "#ffc05c", "Non-Hostile" = "#FFF159", "Sleep-Friendly" = "lightgreen", "Unknown" = "grey70"),
        layer.name = "Service area") +
  
  mapview(inex_benches, zcol = "is_inclusive",
        col.regions = c("hostile" = "#C7701A", "non-hostile" = "yellow3", "sleep-friendly" = "green4"),
        na.color = "grey60",
        cex = 3, layer.name = "Benches")



```


# IDK anymore


```{r, voronoi summary}

# IDK what this is trying to do

#cat("=== All benches — service area (m²) ===\n")
#summary(voronoi_all$area_m2)

#cat("\n=== Inclusive benches only — service area (m²) ===\n")
#summary(voronoi_in$area_m2)

#ratio <- mean(voronoi_in$area_m2) / mean(voronoi_all$area_m2)
#cat(sprintf("\nMean service area is %.1fx larger when restricted to inclusive benches.\n", ratio))

```


## Demographic Overlay (AR) -> look into this

An idea for using dkstat data which is a part of the opengov dk intitative: https://github.com/rOpenGov/dkstat 
We could use IDs to identify different demographics spread through aarhus - not sure if an ID for e.g., 'homeless' or socialt udsatte exists or one for elderly people exists but it might be a nice (maybe) way to get an understanding of how the presence and distribution of benches say something about how we care (or not) about various demographics of people.


## HELP for Paper stuff_to be deleted

```{r data}

print(unique(st_geometry_type(raw_streetNoise)))

```

```{r citation}

pacman::p_load(attachment)
attachment::att_from_rmd("MAIN.Rmd") # packages used

packageVersion("mapview")
citation(package="mapview")


```
