---
title: "Road_Noise_Pollution_Data_Download"
course: "Cultural Datascience: Spatial Analytics at AU"
authors: "Aiswarya Roy & Mie Norre Engemann"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
---

_Script Description:_  
The current script was used to download the road noise pollution data for Denmark for city roads during the night (22-07) (i.e., Lnight) at 1.5 meters height from 2022. The noise pollution data for each basemap (aarhus municipality and two based on Aarhus ringvej and Aarhus ringgade) is then saved to the data/ folder. This data is what can be found in the GitHub repository. The original datafile is too big, but can be found via this link to Miljøstyrelsens MiljøGIS (i.e., the Danish Environmental Protection Agency's GIS): https://mst.dk/erhverv/tilskud-miljoeviden-og-data/data-og-databaser/miljoegis-data-om-natur-og-miljoe-paa-webkort/hent-data-udstillet-paa-miljoegis.


_Script:_

```{r setup, include=FALSE}
## Global options
library(pacman)

pacman::p_load(rmdformats, knitr)

options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


## Loading Libraries

```{r load_libraries, message = FALSE, warning=FALSE}

library(pacman)

pacman::p_load(sf,
               tidyverse,
               osmdata,
               ggplot2,
               geodata,
               dplyr,
               geopackage,
               mapview)


```


## Data Preparation (ME)

Loading Data, CRS conversion etc.

1) Base Map of Aarhus
```{r download_municipality_data}

## BASEMAP;   inspiration from class scripts
workingDirectory_path <- getwd()

# get municipality data
municipalityData_downloaded <- geodata::gadm(country="Denmark", level=2, path=file.path(workingDirectory_path, "data"), version="3.6") # data from the GADM database and save to data/gadm/

municipalityData_sf <- st_as_sf(municipalityData_downloaded)


```

```{r prepare_municipality_base_map}

## BASEMAP;   inspiration from class scripts

# preparing the data for this project
municipalityData_sf$NAME_2[31] <- "Aarhus" # from århus to aarhus

municipalityData_sf <- subset(municipalityData_sf, select = -c(GID_1, NL_NAME_1, GID_2, VARNAME_2, NL_NAME_2, CC_2)) # removing columns not used


# filter to only extract Aarhus
aarhus_municipality <- filter(municipalityData_sf, NAME_2 == "Aarhus")


# save CRS
baseMap_crs = st_crs(aarhus_municipality) # longlat, WGS84



```


1.1) Ringvejen / Ringgaden (other option for basemap)
```{r}

# load the shapefile for noise pollution from streets during the night at 1.5 meters height
aarhus_ring <- read_sf("data/ringVejGade/ringVejGade.geojson") # custom polygons from geojson.io powered by Mapbox

aarhus_ringvej <- filter(aarhus_ring, Ringvejen == 1)
aarhus_ringgade <- filter(aarhus_ring, Ringgaden == 1)


```



2) Noise Pollution Data

```{r prepare_noise_pollution_data}

# load the shapefile for noise pollution from streets during the night at 1.5 meters height
raw_streetNoise <- st_read("data/noisePollution/dk_2022_vej_nat_1_5m.shp") # downloaded from Miljøstyrelsen at: https://mst.dk/erhverv/tilskud-miljoeviden-og-data/data-og-databaser/miljoegis-data-om-natur-og-miljoe-paa-webkort/hent-data-udstillet-paa-miljoegis


# convert CRS to WGS84 (4326)
raw_streetNoise <- st_transform(raw_streetNoise, baseMap_crs)

# areas to have as basemaps
basemapOptions_aarhus <- c("municipality", "ringvej","ringgade")

# ensure st is valid for the oise data
raw_streetNoise <- st_make_valid(raw_streetNoise)

for (currentBasemap_name in basemapOptions_aarhus){
  # current basemap
  if (currentBasemap_name == "municipality"){
    baseMap <- aarhus_municipality
  } else if (currentBasemap_name == "ringvej"){
    baseMap <- aarhus_ringvej
  } else if (currentBasemap_name == "ringgade"){
    baseMap <- aarhus_ringgade
  } else{
    stop("The 'basemapOptions_aarhus' variable must only consist of: municipality, ringvej and ringgade")
  }
  
  # make st valid
  baseMap <- st_make_valid(baseMap)
  
  # only keep the data within the current basemap (Aarhus municipality, ringvejen or ringgaden)
  aarhus_night_streetNoise <- st_intersection(raw_streetNoise, baseMap)
  
  noise_file <- sprintf("data/aarhus_night_streetNoise_%s.gpkg", currentBasemap_name)
  st_write(aarhus_night_streetNoise, noise_file, delete_dsn = TRUE)
  
}


```


